Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-02-05 paper.tex
minor edits; briefly explain unification

The Helmholtz machine~\cite{dayan1995helmholtz} introduced the idea of alternating between
learning a generative model and training a recognition model on samples from the generative model.
We apply this same idea to the problem of program induction,
and so we call our algorithm the \system.
Figure~\ref{feeding} diagrams how these three steps feed off of each other.
\begin{figure}\centering
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](synthesis) at (6,4) {Search for \\programs: $p$};
    \node[align = center](DSL) at (3,1) {DSL: $\mathcal{D}$};
    \node[align = center](recognitionModel) at (9,1) {Recognition \\model: $q$};

@@

We frame this problem as maximum a posteriori (MAP) inference in the generative model diagrammed in Fig.~\ref{graphicalModel}. We wish to maximize the MAP probability of $\mathcal{D}$:
\begin{equation*}
  \probability[\mathcal{D}|X]\propto \probability[\mathcal{D}] \int \mathrm{d}\theta P(\theta|\mathcal{D})\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{equation*}
In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
Conventional natural-language processing (NLP) approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
\begin{align}
  A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
&+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
  \end{align}

@@

We maximize $J$ by alternating maximization w.r.t. the DSL and the frontiers:
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $J$ w.r.t. the frontiers.} Here we
want to find new programs to add to  the frontiers so that $J$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.
\\\noindent \textbf{DSL Induction: Maxing $J$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed and so we can evaluate $J$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $J$.

Searching for programs is extremely difficult because
of how large the search space is. We ease the difficulty of the search by learning a neural recognition model:

\textbf{Neural recognition model: tractably maxing $J$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large.  With $q$ in hand we can find programs for
frontier $\mathcal{F}_x$ by searching for programs maximizing
$q(p|x)$.
The network $q$ exploits the structure of the DSL $\mathcal{D}$:
rather than directly predicting a distribution over $p$ conditioned on $x$,
it predicts a weight vector, $\theta^{(x)}$, and we define $q(p|x)\triangleq \probability[p|\mathcal{D},\theta = q(x)]$.
This approach implements an amortized
inference scheme~\cite{le2016inference,ritchie2016deep} for the generative model in
Fig.~\ref{graphicalModel}.

@@

To make this concrete we need to define what programs actually are and
what form $\probability[p |\mathcal{D},\theta]$ takes.
In this work, we represent programs as $\lambda$-calculus expressions.
$\lambda$-calculus is a formalism for expressing functional programs
that closely resembles Lisp.
$\lambda$-calculus includes variables, function application, and the ability to create new functions.
Throughout this paper we will write $\lambda$-calculus expressions in Lisp syntax.
Our programs are all strongly typed.
We use the Hindley-Milner polymorphic typing system~\cite{pierce} which is used in functional languages like OCaml.
Variables that range over types are always written using lowercase Greek letters
and we write $\alpha\to \beta$ to mean a function that takes as input type $\alpha$
and returns something of type $\beta$.
We use the notation $p:\tau$ to mean that the $\lambda$-calculus expression $p$
has the type $\tau$.
For example, to describe the type of the identity function
we would say \texttt{(lambda (x) x)}$:\alpha\to \alpha$.
With this notation in hand we can now define DSLs:

@@

\section{Program Representation}
We choose to represent programs using $\lambda$-calculus~\cite{pierce}.
A $\lambda$-calculus expression is either:
\\\noindent A \emph{primitive}, like the number 5 or
  the function \texttt{sum}.
\\\noindent A \emph{variable}, like $x$, $y$, $z$
\\\noindent A $\lambda$\emph{-abstraction}, which creates a new function. $\lambda$-abstractions have a variable and a body. The body is a $\lambda$-calculus expression. Abstractions are written as $\lambda \text{var}. \text{body}$.
\\\noindent An \emph{application} of a function to an argument. Both the function and the argument are $\lambda$-calculus expressions. The application of the function $f$ to the argument $x$ is written as $f\; x$.

>>>>

The Helmholtz machine~\cite{dayan1995helmholtz} introduced the idea of alternating between
learning a generative model and training a recognition model on samples from the generative model.
We apply this idea to program induction, motivating why we call our algorithm the \system.
Figure~\ref{feeding} diagrams how these three steps feed off of each other.
\begin{figure}\centering
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](synthesis) at (6,4) {Search for \\programs: $p$};
    \node[align = center](DSL) at (3,1) {DSL: $\mathcal{D}$};
    \node[align = center](recognitionModel) at (9,1) {Recognition \\model: $q$};

@@

We frame this problem as maximum a posteriori (MAP) inference in the generative model diagrammed in Figure~\ref{graphicalModel}. We wish to maximize the posterior probability of $\mathcal{D}$:
\begin{equation*}
  \probability[\mathcal{D}|X]\propto \probability[\mathcal{D}] \int \mathrm{d}\theta P(\theta|\mathcal{D})\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{equation*}
In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
Conventional natural-language processing (NLP) approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
\begin{align}
  A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
&+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
  \end{align}

@@

We maximize $J$ by alternating maximization w.r.t. the frontiers and the DSL:
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $J$ w.r.t. the frontiers.} Here we
want to find new programs to add to  the frontiers so that $J$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.
\\\noindent \textbf{DSL Induction: Maxing $J$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed so we can evaluate $J$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $J$.

Searching for programs is extremely difficult because
of the large search space. We ease this difficulty by learning a neural recognition model:

\textbf{Neural recognition model: tractably maxing $J$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large.  With $q$ in hand we can find programs for
the frontier of a task $\mathcal{F}_x$ by searching for programs that maximize
$q(p|x)$.
The network $q$ exploits the structure of the DSL $\mathcal{D}$:
rather than directly predicting a distribution over $p$ conditioned on $x$,
it predicts a weight vector, $\theta^{(x)}$, and we define $q(p|x)\triangleq \probability[p|\mathcal{D},\theta = q(x)]$.
This approach implements an amortized
inference scheme~\cite{le2016inference,ritchie2016deep} for the generative model in
Figure~\ref{graphicalModel}.

@@

To make this concrete we need to define what programs actually are and
what form $\probability[p |\mathcal{D},\theta]$ takes.
We represent programs as $\lambda$-calculus expressions.
$\lambda$-calculus is a formalism for expressing functional programs
that closely resembles the Lisp programming language.
$\lambda$-calculus includes variables, function application, and the ability to create new functions.
Throughout this paper we will write $\lambda$-calculus expressions in Lisp syntax.
Our programs are all strongly typed.
We use the Hindley-Milner polymorphic typing system~\cite{pierce} which is used in functional languages like OCaml.
Variables that range over types are always written using lowercase Greek letters
and we write $\alpha\to \beta$ to mean a function that takes as input type $\alpha$
and returns something of type $\beta$.
We use the notation $p:\tau$ to mean that the $\lambda$-calculus expression $p$
has the type $\tau$.
For example, to describe the type of the identity function
we would say \texttt{(lambda (x) x)}$:\alpha\to \alpha$.
We say a type $\alpha$ \emph{unifies} with $\tau$ if every expression
$p:\alpha$ also satisfies $p:\tau$. Furthermore, the act of \emph{unifying}
a type $\alpha$ with $\tau$ is to introduce constraints on the type
variables of $\alpha$ to ensure that $\alpha$ unifies with $\tau$.
See Section~\ref{programrepresentation} for more detail on program representation.
With this notation in hand we can now define DSLs:

@@

\section{Program Representation}\label{programrepresentation}
We choose to represent programs using $\lambda$-calculus~\cite{pierce}.
A $\lambda$-calculus expression is either:
\\\noindent A \emph{primitive}, like the number 5 or
  the function \texttt{sum}.
\\\noindent A \emph{variable}, like $x$, $y$, $z$
\\\noindent A $\lambda$\emph{-abstraction}, which creates a new function. $\lambda$-abstractions have a variable and a body. The body is a $\lambda$-calculus expression. Abstractions are written as $\lambda \text{var}. \text{body}$.
\\\noindent An \emph{application} of a function to an argument. Both the function and the argument are $\lambda$-calculus expressions. The application of the function $f$ to the argument $x$ is written as $f\; x$.
